---
title: "COVID-19 on Diferent Aspects from Toronto Cases"
author: "Zhendong Zhang"
date: 2020-09-25
abstract: "In this paper I will explore the data of COVID-19 cases from Toronto, and analyzing it in different aspects such as gender groups, age groups, fatal and resolved cases. Also, I will develop a model to pursue deeper influences from the variables and to know the level of influences by each variable."
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
#install.packages("opendatatoronto")
library(opendatatoronto)
library(tidyverse)

```


```{r download}

resou <- list_package_resources("https://open.toronto.ca/dataset/covid-19-cases-in-toronto/")
covdat <- get_resource(resou)
str(covdat)
variable.names(covdat)

```

The dataset(covdat) is total reported cases of COVID-19 in Toronto\

Episode Date is an estimate of symptom onset date in a case. It is calculated hierarchically based
on either the date of symptom onset, specimen collection/test date, or the date reported to
public health.\

Reported Date is the date the case was reported to public health.\

# Introduction

The major event in the year 2020 is surely the outbreak of COVID-19. It has significantly changed our life. In this paper, I will explore the influence of COVID-19 on different groups of people in Toronto. And also the effectiveness of the response of Government. This paper is mainly based on the confirmed cases` data from Toronto.\

I mainly compared the data from each group of people such as male, female, resolved cases and fatal cases. Then I tried to find out the characteristics in each group and found out the possible reasons behind these data. The major found is that most fatal cases in the Toronto area are elder people, and this disease has indiscriminate influences across different genders.\

First, I traced the curve of the outbreak from Jan 2020 until now and the distribution of cases on each age group in Toronto. Then I analyzed the sources of infections on these cases, and I trace the closed contacted cases curve to see if the policy is effective. In the end, I mainly focus on the fatal and resolved cases and tried to analyze the characteristics behind these two groups. In the future I will keep monitoring the cases in Toronto, seeing if there is another event worth analyzing. This paper`s weakness is that I omit lots of cases that the data related are not available so we cannot trace them, but there might be a major influence in these cases, which could significantly influence our conclusion. Also, the data sample is too small since we only have Toronto data. In the end, I developed a tree-based model on predicting the possibilities of outcomes for a confirmed case.\

My source code is here: https://github.com/Craymate/STA304-2020/blob/master/content/post/2020-09-20-research-of-covid-19-on-diferent-aspects-from-toronto-cases.Rmd


# <span style="color:purple">For Overall Cases:</span>

```{r infect-month, warning=FALSE, echo=FALSE, results="hide", fig.align='center'}

covdat <- filter(covdat, Classification == "CONFIRMED")
#Mainly focus on the confirmed cases

covdat <- mutate(covdat, month = strftime(as.Date(`Reported Date`),"%m"))

ggplot(covdat, aes(x = month)) + geom_histogram(stat="count") + labs(x = "Months", y = "Number of Confirmed Cases", title = "Cases Distribution Across Months") + theme(plot.title = element_text(hjust = 0.5))

```

The peak is in April, and started to decrease in May. There was a boost from March to April. Then it has been decreased to the level in March. But now it shows a sign of increasing again.

```{r infect-age, warning=FALSE, echo=FALSE, results="hide", fig.align='center'}

ggplot(covdat, aes(x = `Age Group`)) + geom_bar(stat="count") + labs(x = "Age Groups", y = "Number of Confirmed Cases", title = "Cases Distribution Across Age Groups") + theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5), plot.title = element_text(hjust = 0.5))

```

The infection between age groups is nearly normal distributed.

```{r infect-gender, echo=FALSE, results="hide", fig.align='center'}

total_gender <- count(group_by(covdat, `Client Gender`))
total_gender <- mutate(total_gender, Class = "Total")

ggplot(covdat, aes(x = `Client Gender`)) + geom_bar(stat="count") + labs(x = "Genders", y = "Number of Cases", title = "Gender Group Distribution in Overall Cases") + theme(plot.title = element_text(hjust = 0.5))

```

Different gender groups have slightly differences in infected amount.

```{r source-plots, echo=FALSE, results="hide", fig.align='center'}

ggplot(covdat, aes(x = `Source of Infection`)) + geom_bar(stat="count") + labs(title = "Infection Sources Distribution") + theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5), plot.title = element_text(hjust = 0.5))
#The highest source is NA.

```

There are many cases that they are NA-Outbreak Associated, which means that we don`t know the source of infection on these cases.

```{r filtered, echo=FALSE, results="hide", fig.align='center'}

sources <- filter(covdat, `Source of Infection` != "N/A - Outbreak associated")

ggplot(sources, aes(x = `Source of Infection`)) + geom_bar(stat="count") + labs(title = "Filtered Infection Sources Distribution") + theme(plot.title = element_text(hjust = 0.5))

```


The highest source of infection is close contact, which is significant higher than others.
And the second highest is the community.

```{r covdat-contact, echo=FALSE, results="hide", fig.align='center'}

contact <- filter(sources, `Source of Infection` == "Close contact")
#This is only the cases infected by close contact

contact_case <- summarise(group_by(contact, month), Class = "Closed Contact", count=n(), .groups = 'drop')
total_case <- summarise(group_by(covdat, month), Class = "Total", count=n(), .groups = 'drop')
cases <- rbind(contact_case, total_case)
ggplot(cases, aes(x=month, y=count, fill = Class)) + geom_bar(stat="identity", position="dodge") + labs(x = "Months", y = "Number of Cases", title = "Total Cases vs. Confirmed Closed Contact Cases") + theme(plot.title = element_text(hjust = 0.5))

```

Since April 1st, there was an emergency response from Ontario Government. But the close contact confirmed cases increases to the peak in May. Then this policy starts to take effect in June, and making significant effect in the following three months until now. In September , it shows a sign of increasing with the number of close contact cases is also increased.

```{r correlation}
full_contact <- c(0,0,contact_case$count)
cor.test(full_contact, total_case$count)
```

Low P-value from result of cor.test() means total cases is related to close contact cases.
The sample estimates means that the total cases are estimated 88.1% related by closed contact cases.
This means social distancing is expected to be a right strategy on this outbreak.

```{r total-outcome, echo=FALSE}

outcome <- count(group_by(covdat, Outcome))
pie(outcome$n, labels=outcome$Outcome, main = "Outcome Distribution")

```

The majority outcomes are resolved.


# <span style="color:red">For Fatal Cases:</span>

```{r fata-age, echo=FALSE, results="hide", fig.align='center'}

#For the fatal cases
fata <- filter(covdat, covdat$Outcome == "FATAL")
ggplot(fata, aes(x = `Age Group`)) + geom_bar(stat="count") + labs(x = "Age Groups", y = "Number of Cases", title = "Age Group Distribution in Fatal Cases") + theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5), plot.title = element_text(hjust = 0.5))

```

Most of cases are elder people. This is reasonable because they generally have weaker health conditions. As the age increases, the fatal cases also increase.

```{r fata-gender, echo=FALSE, results="hide", fig.align='center'}

fata_gender <- count(group_by(fata, `Client Gender`))
fata_gender <- mutate(fata_gender, Class = "Fatal")

gender_fata_total <- rbind(total_gender, fata_gender)

ggplot(gender_fata_total, aes(x=`Client Gender`, y=n, fill = Class)) + geom_bar(stat="identity", position="dodge") + labs(x = "Months", y = "Number of Cases", title = "Total Cases vs. Fatal Cases on Different Gender") + theme(plot.title = element_text(hjust = 0.5))

```

Different gender groups have slightly differences on the fatal outcomes

```{r fata-gender-percent, echo=FALSE, fig.align='center'}

total_filtered_gender <- filter(total_gender, `Client Gender` != "TRANSGENDER", `Client Gender` != "OTHER")
gender_fata_ratio <- data.frame(Gender = fata_gender$`Client Gender`, Ratio = fata_gender$n / total_filtered_gender$n)

ggplot(gender_fata_ratio, aes(x=Gender, y=Ratio)) + geom_bar(stat="identity") + labs(x = "Gender", y = "Fatal Rate", title = "Fatal Rate Among Different Genders") + theme(plot.title = element_text(hjust = 0.5))
gender_fata_ratio

```
(The rates are calculated by the ratio between fatal and total cases from each groups.)\
Gender makes no influence between Male and Female. Unknown`s fatal rate is high, because the total amount of samples from Unknown is small.

```{r fata-ratio}

fata_mean <- mean(gender_fata_ratio$Ratio)
fata_mean

```

There is an approximately 9.6% <span style="color:red">Fatal Rate</span>.


```{r fata-hosip, echo=FALSE, results="hide", fig.align='center'}

summ1 <- summarise(fata, Total = count(fata),
          Ever_Hospitalized = count(filter(fata, `Ever Hospitalized` == "Yes")), 
          Ever_in_ICU = count(filter(fata, `Ever in ICU` == "Yes")),
          Ever_Intubated = count(filter(fata, `Ever Intubated` == "Yes")))

barplot(unlist(summ1), names.arg=c("Total", "Ever Hospitalized", "Ever in ICU", "Ever Intubated"), main = "Fatal Cases` History of Inpatient Visits")

```

Nearly half of fatal cases have ever hospitalized.

```{r percent}

percent <- (summ1$Ever_Hospitalized + summ1$Ever_in_ICU + summ1$Ever_Intubated) * 100/summ1$Total
percent

```

And about 71% of total fatal cases got treated. This indicates that the shortage of hospital resources to general extent.


# <span style="color:blue">For Resolved Cases:</span>

```{r resol-age, echo=FALSE, results="hide", fig.align='center'}

#For the resolved cases
resol <- filter(covdat, Outcome == "RESOLVED")
ggplot(resol, aes(x = `Age Group`)) + geom_bar(stat="count") + labs(x = "Age Groups", y = "Number of Cases", title = "Age Group Distribution in Resolved Cases") + theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5), plot.title = element_text(hjust = 0.5))

```

Younger people seems more likely to recover from this.

```{r resol-gender, echo=FALSE, results="hide", fig.align='center'}

resol_gender <- count(group_by(resol, `Client Gender`))
resol_gender <- mutate(resol_gender, Class = "Resolved")

gender_res_total <- rbind(total_gender, resol_gender)

ggplot(gender_res_total, aes(x=`Client Gender`, y=n, fill = Class)) + geom_bar(stat="identity", position="dodge") + labs(x = "Months", y = "Number of Cases", title = "Total Cases vs. Resolved Cases on Different Gender") + theme(plot.title = element_text(hjust = 0.5))

```

Different gender groups also have slightly differences on the resolved outcomes

```{r resol-gender-percent, echo=FALSE, results="hide", fig.align='center'}

gender_resol_ratio <- data.frame(Gender = resol_gender$`Client Gender`, Ratio = resol_gender$n / total_gender$n)

ggplot(gender_resol_ratio, aes(x=Gender, y=Ratio)) + geom_bar(stat="identity") + labs(x = "Gender", y = "Resolve Rate", title = "Resolved Rate Among Different Genders") + theme(plot.title = element_text(hjust = 0.5))
gender_resol_ratio

```

(The rates are calculated by the ratio between resolved and total cases from each groups.)\
Gender nearly makes no influence among different gender groups.


```{r resol-ratio}

resol_mean <- mean(gender_resol_ratio$Ratio)
resol_mean

```

There is an approximately 90.5% <span style="color:blue">Resolved Rate</span>.

```{r resol-hosip, echo=FALSE, results="hide", fig.align='center'}

summ <- summarise(resol, Total = count(resol),
          Ever_Hospitalized = count(filter(resol, `Ever Hospitalized` == "Yes")), 
          Ever_in_ICU = count(filter(resol, `Ever in ICU` == "Yes")),
          Ever_Intubated = count(filter(resol, `Ever Intubated` == "Yes")))

barplot(unlist(summ), names.arg=c("Total", "Ever Hospitalized", "Ever in ICU", "Ever Intubated"), main = "Resolved Cases` History of Inpatient Visits")

```

We can see that in most resolved cases, people can recover without even being hospitalized.
So immune system is making a great effect in this outbreak.


# <span style="color:green">Classification Tree Model:</span>

This is a classification tree based model on predicting the chance you can recover from COVID-19.\
I build this model by using 60% of total cases for training, and test the model on the remaining 40%.\
I cleaned the active cases and made data binomial. This means in the result section, 1 is for Resolved, and 0 is for Fatal.\
I used "Age Group", "Ever Hospitalized", "Ever in ICU", "Ever Intubated", "Neighbourhood Name", "FSA" as my variables. I do not use gender as a variable becuase it has no ifluence on the outcomes, as well as source of infections.\

<span style="color:red">Attention:</span> This model is only for study, the result does`t represent real outcome in reality.\
<span style="color:red">Pre-condition:</span> The input cases must be confirmed.


```{r install, warning=FALSE, include=FALSE}

#install.packages("rpart")
#install.packages("caret")
#install.packages("rpart.plot")

library(rpart)
library(caret)
library(rpart.plot)

```


```{r tree, echo=FALSE, results="hide"}

set.seed(731)

clean <- filter(covdat, Outcome != "ACTIVE")
covdat1 <- mutate(clean, Result = ifelse(Outcome == "RESOLVED", 1, 0) )

#Training Dataset
train <- covdat1[sample(nrow(covdat1), size=0.6 * nrow(covdat1), replace=FALSE),]
train <- select(train, Result, `Age Group`, `Ever Hospitalized`, `Ever in ICU`, `Ever Intubated`, `Neighbourhood Name`, `FSA`)

#Testing Dataset
test <- covdat1[!rownames(covdat1) %in% rownames(train),]
result <- test$Result
test <- select(test,`Age Group`, `Ever Hospitalized`, `Ever in ICU`, `Ever Intubated`, `Neighbourhood Name`, `FSA`)

```

```{r input}

test[1,]

```

The input vector is as above.
FSA stands for the patient`s first three letter of postal code.

```{r data-ratio}

nrow(train)
nrow(test)

```

There are total 9193 training cases.
There are total 6129 testing cases.
\


```{r shorten-lab, echo=FALSE, results="hide"}

split.fun <- function(x, labs, digits, varlen, faclen)
{
    # replace commas with spaces (needed for strwrap)
    labs <- gsub(",", " ", labs)
    for(i in 1:length(labs)) {
        # split labs[i] into multiple lines
        labs[i] <- paste(strwrap(labs[i], width=25), collapse="\n")
    }
    labs
}

```

```{r test-mod, warning=FALSE, echo=FALSE, results="hide"}

#Fitting Model
tree <- rpart(Result~`Age Group` + `Ever Hospitalized` + `Ever in ICU` + `Ever Intubated` + `Neighbourhood Name` + `FSA`, data=covdat1, method = "class")
pdf("Tree Model.pdf") 
rpart.plot(tree, split.fun=split.fun, main="Tree Model")
dev.off()

rpart.plot(tree, split.fun=split.fun, main="Tree Model")

```

Here is the Tree Model Plot: https://github.com/Craymate/STA304-2020/blob/master/content/post/Tree%20Model.pdf\

```{r test-pred}

Pred <- predict(tree, test, type = "class")
Reference <- as.matrix(result)
confusionMatrix(table(Pred, result))

```

The rate of accuracy is approximately 97.7%.\
And there are 105 cases actually fatal but predicted resolved.\
And there are 36 cases actually resolved but predicted fatal.\


```{r try, include=FALSE}

try1 <- test[0,]
try1[1,] <- as.list(c("20 to 29 Years", "No", "No", "No", "Little Portugal", "M6K"))
try1

try2 <- test[0,]
try2[1,] <- as.list(c("80 to 89 Years", "No", "No", "No", "Little Portugal", "M6K"))

try3 <- test[0,]
try3[1,] <- as.list(c("80 to 89 Years", "Yes", "Yes", "Yes", "Victoria Village", "M4A"))

```

<span style="color:green">Here we can try to create cases to see their outcomes.</span>\

The first case is in "20 to 29 years" age group.And not ever in ICU, Hospitalized, Intubated.\
Neighbourhood is "Little Portugal"\
FSA is "M6K"\

The second case is in "80 to 89 Years" age group. And the rest are same\

The third case is in "80 to 89 Year" And have ever in ICU, Hospitalized, Intubated.\
Neighbourhood is "Victoria Village"\
FSA is "M4A"\
\


```{r try1-pred}

predict(tree, try1)

```

Then we can plug the first case into the model.\
The outcome is 0.75% to be fatal and 99.25% to be resolved.\

```{r try2-pred}

predict(tree, try2)

```

The second case which has an older age, has 21.3% chance to be fatal, and 78.7% to be resolved.\
In this case, we can identify the age influences on the outcome.\

```{r try3-pred}

predict(tree, try3)

```

The third case has the same age with the second one. But changed the location and the medical history.\
The outcome becomes 88.24% to be fatal and 11.76% to be resolved.\

# Discussion

COVID-19 is making no difference on different gender groups, but age groups. Elder people cases are more likely to be fatal based on Toronto data. And recovery of this disease is heavily based on immune system. Also, every age group could have equally chance to get infected.\

Despite the unknown source of infection cases, the major source is close contact. The government published a law in April to enforce social distancing. This law is making its effect in June, and it is effective. The level of confirmed cases in Toronto in Sept shows the sign of increasing now.\

We can identify the influences of medical history from COVID-19 in the tree model, as well as the influences by age. With age increasing, the possibility of fatal is increasing. And if a patient has ever been to hospital, ICU or intubated, the possibility also has different levels of increases.\

# Weakness

In this paper, I omit the NA sources of infection, but it has the highest amount of infections, so this might be critical. Also, the data-set is too small to trace the real characteristics of COVID-19, since it only has data from Toronto. In the future, I will use a larger and well-specified data to further my study.\

The data in this spreadsheet are subject to change as public health investigations into reported cases and continuous quality improvement initiatives are ongoing, and additional cases continue to be reported. The data will be completely refreshed and overwritten on a weekly basis.\

This tree-based model does not have much precision on predicting even though the testing result is good. Because this model is lack of advanced algorithms and sample data, we can only identify some variable influences by it.\

# References

- https://www.toronto.ca/home/covid-19/covid-19-protect-yourself-others/covid-19-reduce-virus-spread/
- https://open.toronto.ca/dataset/covid-19-cases-in-toronto/
- Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686,
  https://doi.org/10.21105/joss.01686
- Sharla Gelfand (2020). opendatatoronto: Access the City of Toronto Open Data Portal.
  https://sharlagelfand.github.io/opendatatoronto/,
  https://github.com/sharlagelfand/opendatatoronto/.
- Terry Therneau and Beth Atkinson (2019). rpart: Recursive Partitioning and Regression Trees. R package version
  4.1-15. https://CRAN.R-project.org/package=rpart
- Stephen Milborrow (2020). rpart.plot: Plot 'rpart' Models: An Enhanced Version of 'plot.rpart'. R package
  version 3.0.9. https://CRAN.R-project.org/package=rpart.plot

